---
title: "R Notebook"
output: html_notebook
---
library(tidyverse)
library(dplyr)
library(ggplot2)


Missing Persons Exploration 
read in the NCIC text file of missing persons by Race, 2007-2019
```{r}
ncic <- read_csv("data/ncic_missing.txt")
view(ncic)
```
drop unnecessary columns for now 
```{r}
ncic <- ncic %>% 
  select(-"Total", -"Female_ratio_Pop", -"AIAN_ratio")
```
convert column to date (Year) and confirm others are numeric (Races)
```{r}
sapply(ncic,class)
```
Run numbers by Race and Year
```{r}
ncic_stacked <- ncic %>%   
  gather("AIAN", "Asian", "Black", "Unk", "White", key = Race, value = Count)
ncic_stacked

```
convert to csv to use in app 
```{r}
write.csv(ncic_stacked, "data/ncic_stacked.csv")
```

```{r}
ncic_stacked %>% 
  ggplot(aes(x = Year, y = Count, fill = Race)) +
  geom_area()
```
Compute using proportions 
```{r}
# Compute percentages with dplyr
ncic_proport <- ncic_stacked  %>%
  group_by(Year, Race) %>%
  summarise(n = sum(Count)) %>%
  mutate(Percentage = n / sum(n))

# Plot
ggplot(ncic_proport, aes(x=Year, y=Percentage, fill=Race)) + 
    geom_area(alpha=0.6 , size=1, colour="black")
```
```{r}
view(ncic_proport)
```
```{r}
ncic_proport %>% 
  ggplot(aes(x = Year, y = Percentage, color = Race))+ 
  geom_line()
```
```{r}
write_csv(ncic_proport, "data/ncic_proport.csv")
```
Read in WISQARS pop by race report. Combine with ncic_stacked in order to estimate missing rate by Race  
```{r}
mf_pop <- read_csv("data/WISQARS_race_pop_allsex_2007_2018.csv")

```
```{r}
mf_pop <- mf_pop %>% 
  select("Race", "Year", "2013 Urbanization (Collapsed) Classification", "Population")
```
rename 2013 Urbanization...

```{r}
mf_pop <- mf_pop %>% 
  rename(Metro_type = "2013 Urbanization (Collapsed) Classification")
```
```{r}
#view(mf_pop)
```

```{r}
mf_pop$Metro_type[is.na(mf_pop$Metro_type)] <- "All Metros"
```
```{r}
mf_pop$Race[mf_pop$Race == "Am Indian/AK Native"] <-  "AIAN"
```
```{r}
mf_pop$Race[mf_pop$Race == "Asian/Pac Islander"] <-  "Asian"
```
```{r}
mf_pop_filter <- mf_pop %>% 
  filter(Metro_type == "All Metros")
```
arrange Race in same order as the ncic_stacked dataset
```{r}
mf_pop_filter %>% 
  arrange(Year) %>% 
  select(-"Metro_type")
```
```{r}
#write.csv(mf_pop_filter, "data/mf_pop_filter.csv", row.names = TRUE)
```

```{r}
#ncic_pop <- inner_join(mf_pop_filter, ncic_stacked, by = "Year")
```
```{r}
#view(ncic_pop)
```

```{r}
ncic_proport2 <- ncic_proport %>% 
  filter(Race != "Unk")
```
```{r}
view(ncic_proport2)
```
```{r}
mf_pop_filter2 <- read_csv("data/mf_pop_filter_w_2019.csv")
```

```{r}
view(mf_pop_filter2)
```
```{r}
mf_pop_filter2 %>% 
  select(-"Metro_type", -"X1")
```
```{r}
mf_pop_filter2 %>% 
  arrange(Year, Race) %>% 
  select("Race", "Year", "Population")
```
```{r}
write.csv(mf_pop_filter2, "data/mf_pop_filter2.csv")
```
```{r}
ncic_pop_proport <- merge(ncic_proport2, mf_pop_filter2)
ncic_pop_proport
```
```{r}
ncic_pop_proport <- ncic_pop_proport %>% 
  select(-"X1", -"Metro_type")
```
```{r}
ncic_pop_proport <- ncic_pop_proport %>% 
  mutate(Population_rate = n/Population)
  
```
```{r}
#view(ncic_pop_proport)
```
```{r}
ncic_pop_proport <- ncic_pop_proport %>% 
  rename(Number_missing = "n")
```
```{r}
ncic_pop_proport %>% 
  ggplot(aes(x = Year, y = Population_rate, fill = Race)) +
  geom_area()
```
```{r}
ncic_pop_proport %>% 
ggplot(aes(x=Year, y=Population_rate, fill=Race)) + 
    geom_area(alpha=0.6 , size=1, colour="black")
```
```{r}
ncic_pop_proport2 <- ncic_pop_proport %>%
  group_by(Year, Race) %>%
  #summarise(n = sum(Count)) %>%
  mutate(Rate100 = Population / sum(Population))

# Plot
ggplot(ncic_proport, aes(x=Year, y=Percentage, fill=Race)) + 
    geom_area(alpha=0.6 , size=1, colour="black")
```

