---
title: "R Notebook"
output: html_notebook
---

library(tidyverse)
library(censusapi)
library(dplyr)
library(tidycensus)
library(ggplot2)


```{r}
#created a text file with the api key, read in
key=read_lines('data/key.txt')
```
```{r}
#create dataset for us census apis
apis <- listCensusApis()
#view(apis)
```
```{r}
#look into metadata for acs1/2019
metadata <- listCensusMetadata(
  name = "acs/acs1/profile",
  vintage = 2019)
```
```{r}
#view(metadata)
```
```{r}
#listCensusMetadata(type = "variables")
```
Pull in General US population data on race, exclude "Hispanic or Latino" ("DP05_0071PE")
```{r}
gen_pop <- getCensus(
  key = key,
  name = "acs/acs1/profile", 
  vintage = 2019,
  vars = c("DP05_0033PE", "DP05_0039PE", "DP05_0038PE", "DP05_0044PE", "DP05_0052PE", "DP05_0037PE", "DP05_0035PE", "DP05_0057PE"),
  #United States
  region = "us:*") 
  #State of TN
  #regionin = 'state:47') 

```
Rename columns
```{r}
gen_pop <- gen_pop %>% 
  rename("Total Pop" = DP05_0033PE, "American Indian and Alaska Native" = DP05_0039PE, "Asian" = DP05_0044PE, "Black or African American" = DP05_0038PE, "Native Hawaiian and Other Pacific Islander" = DP05_0052PE, "White" = DP05_0037PE, "Two or more races" = DP05_0035PE, "Some Other Race" = DP05_0057PE)
```
```{r}
view(gen_pop)
```

convert to test csv file
```{r}
write.csv(gen_pop,"data/us_pop_stats.csv", row.names = TRUE)
```
Read in National Victimization Report, specify columns for Year, Race, Violent Crime, #select_vars(vars = "Year", "V2040A", "V3086", "V4529")
```{r}
natl_crime <- read_tsv("data/37689-0002-Data.tsv")
 
```

```{r}
natl_crime
#natl_crime <- natl_crime %>% 
#select(Year, xx) 
```
Read in and merge the NamUS datasets
```{r}
# miss_AIAN <- read_csv("data/NamUs_Misssing_AIAN_Female.csv")
# view(miss_AIAN)
```
Read in all female missing persons dataset
```{r}
miss_all <- read_csv("data/NamUs_all_fem_missing.csv")

```
```{r}
view(miss_all)
```
Remove unnecessary columns 
```{r}
miss_all_narrow <- miss_all%>% 
  subset(select=-c(`Last Name`, `First Name`, `Missing Age`, City, County, Sex, `Date Modified`))
  
```
convert DLC to date 
```{r}
list(miss_all_narrow)
```
```{r}
class(miss_all_narrow$DLC)
```
```{r}
miss_all_narrow["DLC_new"] <- lapply(miss_all_narrow["DLC"] , as.Date, "%m/%d/%Y" )
```

```{r}
view(miss_all_narrow)
```


```{r}
#miss_all <- miss_all %>% 
  #rename(miss_all, Case_Number = `Case Number`, Date_Last_Contact = DLC, Race = `Race / Ethnicity`)

```

plot all NamUs female missing persons by race and date 
```{r}
#miss_all %>% 
  #ggplot(aes(x = DLC_new, y = Race / Ethnicity))+
  #geom_line()

```
bring in CDC violent deaths dataset
```{r}
viol_deaths <- read.csv("data/WISQARS_all_race_vio_deaths_suic_female_all_metro_by_year.csv")
```
```{r}
print(viol_deaths)
```


