---
title: "R Notebook"
output: html_notebook
---

library(tidyverse)
library(censusapi)
library(dplyr)
library(tidycensus)
library(ggplot2)
library(prettyR)


```{r}
#created a text file with the api key, read in
key=read_lines('data/key.txt')
```
```{r}
#create dataset for us census apis
apis <- listCensusApis()
#view(apis)
```
```{r}
#look into metadata for acs1/2019
metadata <- listCensusMetadata(
  name = "acs/acs1/profile",
  vintage = 2019)
```
```{r}
#view(metadata)
```
```{r}
#listCensusMetadata(type = "variables")
```
Pull in General US population data on race, exclude "Hispanic or Latino" ("DP05_0071PE")
```{r}
gen_pop <- getCensus(
  key = key,
  name = "acs/acs1/profile", 
  vintage = 2019,
  vars = c("DP05_0033PE", "DP05_0039PE", "DP05_0038PE", "DP05_0044PE", "DP05_0052PE", "DP05_0037PE", "DP05_0035PE", "DP05_0057PE"),
  #United States
  region = "us:*") 
  #State of TN
  #regionin = 'state:47') 

```
Rename columns
```{r}
gen_pop <- gen_pop %>% 
  rename("Total Pop" = DP05_0033PE, "American Indian and Alaska Native" = DP05_0039PE, "Asian" = DP05_0044PE, "Black or African American" = DP05_0038PE, "Native Hawaiian and Other Pacific Islander" = DP05_0052PE, "White" = DP05_0037PE, "Two or more races" = DP05_0035PE, "Some Other Race" = DP05_0057PE)
```
```{r}
#view(gen_pop)
```

convert to test csv file
```{r}
write.csv(gen_pop,"data/us_pop_stats.csv", row.names = TRUE)
```
Read in National Victimization Report, specify columns for Year, Race, Violent Crime
```{r}
#natl_crime <- load(file = "data/37689-0003-Data.rda")
#natl_crime <- read_tsv("data/37689-0003-Data.tsv")
```
Keep columns for Year, Current Gender, Race, Crime Type.  Weight using Series_Weight (per BJS)

```{r}
#natl_crime <- natl_crime %>% 
  #select("Year", "V3086", "V3023a", "V3024a", "V4529", "SERIES_WEIGHT")

```
Read in and merge the NamUS missing AIAN dataset and clean 
```{r}
miss_AIAN <- read_csv("data/NamUs_Missing_AIAN_Female.csv")
view(miss_AIAN)
```
```{r}
miss_AIAN <- miss_AIAN%>% 
  subset(select=-c(`Last Name`, `First Name`, `Missing Age`, City, County, Sex, `Date Modified`))
```
```{r}
miss_AIAN["DLC_new"] <- lapply(miss_AIAN["DLC"] , as.Date, "%m/%d/%Y" )
```
```{r}
miss_AIAN <- miss_AIAN %>% 
  rename(Case_Number = "Case Number", Date_Last_Contact = "DLC_new", Race_Ethnic = "Race / Ethnicity")
```
```{r}
miss_AIAN <- miss_AIAN %>% 
  select(-"DLC")
```
```{r}
view(miss_AIAN)
```


Read in all NamUs female missing persons dataset, all races
```{r}
miss_all <- read_csv("data/NamUs_all_fem_missing.csv")

```
```{r}
view(miss_all)
```
Remove unnecessary columns and rename to "miss_all_narrow"
```{r}
miss_all_narrow <- miss_all%>% 
  subset(select=-c(`Last Name`, `First Name`, `Missing Age`, City, County, Sex, `Date Modified`))
  
```
convert DLC to date 
```{r}
list(miss_all_narrow)
```
```{r}
class(miss_all_narrow$DLC)
```
```{r}
miss_all_narrow["DLC_new"] <- lapply(miss_all_narrow["DLC"] , as.Date, "%m/%d/%Y" )
```

```{r}
#view(miss_all_narrow)
```


```{r}
miss_all_narrow <- miss_all_narrow %>% 
  rename(Case_Number = "Case Number", Date_Last_Contact = "DLC_new", Race_Ethnic = "Race / Ethnicity")

```
```{r}
miss_all_narrow <- miss_all_narrow %>% 
  select(-"DLC")
```
remove multi-races from Race_Ethnic column (\,[^,]*$)
```{r}
#miss_all_race <- miss_all_narrow %>% 
 #(str_extract(miss_all_narrow$Race_Ethnic, "\\,[^,]*$"))
```
plot all NamUs female missing persons by race and date 
```{r}
miss_AIAN %>% 
  #mutate(Case_Count = sum(Case_Number)) %>% 
  ggplot(aes(x = Date_Last_Contact))+
  geom_line()
  #scale_y_log10()

```
bring in WISQARS CDC violent deaths datasets, remove unneeded columns, rename remaining columns
```{r}
```
 
Non-metro report, female, all deaths (homicide & suicide), all races 

```{r}
deathnm <- read_csv("data/WISQARS_Vio_Death_Suicide_Women_Race_2001_2018_Nonmetro.csv")
#view(deathnm)

```
```{r}
deathnm <- deathnm %>% 
  select(-"Number of Deaths", -"Crude Rate", -"YPLL", -"Crude YPLL Rate", -"Age-Adjusted YPLL Rate", -"appliedFilters")
```

```{r}
#colnames(deathnm)
```
```{r}
deathnm <- deathnm %>% 
  rename(Race_Ethnic = "Race / Ethnicity", AA_Rate_Nonmetro = "Age Adjusted Rate")

```
```{r}
#view(deathnm)
```
Metro report, female, all deaths (homicide & suicide), all races
```{r}
deathm <- read_csv("data/WISQARS_Vio_Death_Suicide_Women_Race_2001_2018_Metro.csv")
#view(deathm)
```
```{r}
deathm <- deathm %>% 
  select(-"Number of Deaths", -"Crude Rate", -"YPLL", -"Crude YPLL Rate", -"Age-Adjusted YPLL Rate", -"appliedFilters")
```
```{r}
deathm <- deathm %>% 
  rename(Race_Ethnic = "Race / Ethnicity", AA_Rate_Metro = "Age Adjusted Rate")
```
```{r}
#view(deathm)
```

Combined (all) metro report, female, all deaths (homicide & suicide), all races
```{r}
deatham <- read_csv("data/WISQARS_Vio_Death_Suicide_Women_Race_2001_2018_Allmetro.csv")
#view(deatham)
```
```{r}
deatham <- deatham %>% 
  select(-"Number of Deaths", -"Crude Rate", -"YPLL", -"Crude YPLL Rate", -"Age-Adjusted YPLL Rate", -"appliedFilters")
```
```{r}
deatham <- deatham %>% 
  rename(Race_Ethnic = "Race / Ethnicity", AA_Rate_Allmetro = "Age Adjusted Rate")
```
```{r}
#view(deatham)
```

Merge CDC Fatal Deaths, called cdam (combined death all metros) and convert to .csv file
```{r}
cdm <- inner_join(deathm, deatham, by = "Race_Ethnic")
#view(cdm)
```
```{r}
cdam <- inner_join(cdm, deathnm, by = "Race_Ethnic")
view(cdam)
```
```{r}
write.csv(cdam,"data/us_deaths.csv", row.names = TRUE)
```


