---
title: "R Notebook"
output: html_notebook
---

library(tidyverse)
library(censusapi)
library(dplyr)
library(tidycensus)
library(ggplot2)
library(prettyR)


```{r}
#created a text file with the api key, read in
key=read_lines('data/key.txt')
```
```{r}
#create dataset for us census apis
apis <- listCensusApis()
#view(apis)
```
```{r}
#look into metadata for acs1/2019
metadata <- listCensusMetadata(
  name = "acs/acs1/profile",
  vintage = 2019)
```
```{r}
#view(metadata)
```
```{r}
#listCensusMetadata(type = "variables")
```
Pull in General US population data on race, exclude "Hispanic or Latino" ("DP05_0071PE")
```{r}
gen_pop <- getCensus(
  key = key,
  name = "acs/acs1/profile", 
  vintage = 2019,
  vars = c("DP05_0033PE", "DP05_0039PE", "DP05_0038PE", "DP05_0044PE", "DP05_0052PE", "DP05_0037PE", "DP05_0035PE", "DP05_0057PE"),
  #United States
  region = "us:*") 
  #State of TN
  #regionin = 'state:47') 

```
Rename columns
```{r}
gen_pop <- gen_pop %>% 
  rename("Total Pop" = DP05_0033PE, "American Indian and Alaska Native" = DP05_0039PE, "Asian" = DP05_0044PE, "Black or African American" = DP05_0038PE, "Native Hawaiian and Other Pacific Islander" = DP05_0052PE, "White" = DP05_0037PE, "Two or more races" = DP05_0035PE, "Some Other Race" = DP05_0057PE)
```
```{r}
#view(gen_pop)
```

convert to test csv file
```{r}
write.csv(gen_pop,"data/us_pop_stats.csv", row.names = TRUE)
```
Read in National Victimization Report, specify columns for Year, Race, Violent Crime
```{r}
#natl_crime <- load(file = "data/37689-0003-Data.rda")
#natl_crime <- read.table(file = "data/37689-0003-Data.tsv", sep = '\t', header = TRUE)
```
Keep columns for Year, Current Gender, Race, Crime Type.  Weight using Series_Weight (per BJS)

```{r}
#natl_crime <- natl_crime %>% 
  #select("Year", "V3086", "V3023a", "V3024a", "V4529", "SERIES_WEIGHT")

```
#Read in and merge the NamUS missing AIAN dataset and clean 
```{r}
miss_AIAN <- read_csv("data/NamUs_Missing_AIAN_Female.csv")
#view(miss_AIAN)
```
```{r}
miss_AIAN <- miss_AIAN%>% 
  subset(select=-c(`Last Name`, `First Name`, `Missing Age`, City, County, Sex, `Date Modified`))
```
```{r}
miss_AIAN["DLC_new"] <- lapply(miss_AIAN["DLC"] , as.Date, "%m/%d/%Y" )
```
```{r}
miss_AIAN <- miss_AIAN %>% 
  rename(Case_Number = "Case Number", Date_Last_Contact = "DLC_new", Race_Ethnic = "Race / Ethnicity")
```
```{r}
miss_AIAN <- miss_AIAN %>% 
  select(-"DLC")
```
```{r}
#view(miss_AIAN)
```


Read in all NamUs female missing persons dataset, all races
```{r}
miss_all <- read_csv("data/NamUs_all_fem_missing.csv")

```
```{r}
#view(miss_all)
```
Remove unnecessary columns and rename to "miss_all_narrow"
```{r}
miss_all_narrow <- miss_all%>% 
  subset(select=-c(`Last Name`, `First Name`, `Missing Age`, Sex, `Date Modified`))
  
```
convert DLC to date 
```{r}
#list(miss_all_narrow)
```
```{r}
class(miss_all_narrow$DLC)
```
```{r}
miss_all_narrow["DLC_new"] <- lapply(miss_all_narrow["DLC"] , as.Date, "%m/%d/%Y" )
```

```{r}
#view(miss_all_narrow)
```


```{r}
miss_all_narrow <- miss_all_narrow %>% 
  rename(Case_Number = "Case Number", Date_Last_Contact = "DLC_new", Race_Ethnic = "Race / Ethnicity")

```
```{r}
miss_all_narrow <- miss_all_narrow %>% 
  select(-"DLC")
```
remove multi-races from Race_Ethnic column (\,[^,]*$)
```{r}
#miss_all_race <- miss_all_narrow %>% 
 #(str_extract(miss_all_narrow$Race_Ethnic, "\\,[^,]*$"))
```
plot all NamUs female missing persons by race and date 
```{r}
miss_AIAN %>%
  group_by(Date_Last_Contact) %>%
  count(Case_Number, sort = TRUE)
  
```
```{r}
miss_AIAN %>%
  group_by(Date_Last_Contact)
```
```{r}
miss_AIAN %>% 
  group_by(Date_Last_Contact, sort = TRUE) %>% 
  mutate(count = row_number()) %>%   
  mutate(Missing_Count = cumsum(count))
```


```{r}
#ggplot(aes(x = Date_Last_Contact, y = Missing_Number))+
  #geom_line()
  
```
```{r}

```



bring in WISQARS CDC violent deaths datasets, remove unneeded columns, rename remaining columns
```{r}
```
 
Non-metro report, female, all deaths (homicide & suicide), all races 

```{r}
deathnm <- read_csv("data/WISQARS_Vio_Death_Suicide_Women_Race_2001_2018_Nonmetro.csv")
#view(deathnm)

```
```{r}
deathnm <- deathnm %>% 
  select(-"Number of Deaths", -"Crude Rate", -"YPLL", -"Crude YPLL Rate", -"Age-Adjusted YPLL Rate", -"appliedFilters")
```

```{r}
#colnames(deathnm)
```
```{r}
deathnm <- deathnm %>% 
  rename(Race_Ethnic = "Race / Ethnicity", AA_Rate_Nonmetro = "Age Adjusted Rate")

```
```{r}
#view(deathnm)
```
Metro report, female, all deaths (homicide & suicide), all races
```{r}
deathm <- read_csv("data/WISQARS_Vio_Death_Suicide_Women_Race_2001_2018_Metro.csv")
#view(deathm)
```
```{r}
deathm <- deathm %>% 
  select(-"Number of Deaths", -"Crude Rate", -"YPLL", -"Crude YPLL Rate", -"Age-Adjusted YPLL Rate", -"appliedFilters")
```
```{r}
deathm <- deathm %>% 
  rename(Race_Ethnic = "Race / Ethnicity", AA_Rate_Metro = "Age Adjusted Rate")
```
```{r}
#view(deathm)
```

Combined (all) metro report, female, all deaths (homicide & suicide), all races
```{r}
deatham <- read_csv("data/WISQARS_Vio_Death_Suicide_Women_Race_2001_2018_Allmetro.csv")
#view(deatham)
```
```{r}
deatham <- deatham %>% 
  select(-"Number of Deaths", -"Crude Rate", -"YPLL", -"Crude YPLL Rate", -"Age-Adjusted YPLL Rate", -"appliedFilters")
```
```{r}
deatham <- deatham %>% 
  rename(Race_Ethnic = "Race / Ethnicity", AA_Rate_Allmetro = "Age Adjusted Rate")
```
```{r}
#view(deatham)
```

Merge CDC Fatal Deaths, called cdam (combined death all metros) and convert to .csv file
```{r}
cdm <- inner_join(deathm, deatham, by = "Race_Ethnic")
#view(cdm)
```
```{r}
cdam <- inner_join(cdm, deathnm, by = "Race_Ethnic")
#view(cdam)
```
```{r}
write.csv(cdam,"data/us_deaths.csv", row.names = TRUE)
```
```{r}
cdam %>% 
  ggplot(aes(y = AA_Rate_Metro, x = Race_Ethnic))+
  geom_bar(stat = 'identity')+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```
```{r}
cdam %>% 
  ggplot(aes(y = AA_Rate_Nonmetro, x = Race_Ethnic))+
  geom_bar(stat = 'identity')+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```
```{r}
cdam %>% 
  ggplot(aes(y = AA_Rate_Allmetro, x = Race_Ethnic, color = Race_Ethnic))+
  geom_bar(stat = 'identity')+
  ggtitle("Female Violent Deaths by Race, all Metros, 2001-2018")+
  #scale_colour_gradientn(colours=Spectral(4))
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```
Read in Trends report on violent deaths, and compare # of deaths against age-adjusted rates
```{r}
death_trend <- read_csv("data/WISQARS_Trends_all_deaths_all_race_all_metro.csv")
#view(death_trend)
```
```{r}
death_trend %>% 
  ggplot(aes(x = Year, y = `Number of Deaths`, color = `Race / Ethnicity`))+
  geom_line()
```
```{r}
death_trend %>% 
  ggplot(aes(x = Year, y = `Age Adjusted Rate` , color = `Race / Ethnicity`))+
  geom_line()
```
Read in NCIC Missing Rate 
```{r}
ncic_miss <- read_csv("data/ncic_missing.txt")
#view(ncic_miss)
```

```{r}
us_deaths <- read_csv("data/us_deaths.csv")
#view(us_deaths)
```
Read in WISQARS' revised homicide, suicide and both hom/suic, non-hispanic, reports with 2013 collapsed metro indicator classification

```{r}
hom <- read_csv("data/WISQARS_hom_nonhisp_2013_2013uc.csv")
#view(hom)
```
```{r}
suic <- read_csv("data/WISQARS_suic_nonhisp_2013_2013uc.csv")
#view(suic)
```
```{r}
both_death <- read_csv("data/WISQARS_hom_suic_nonhisp_2013uc.csv")
#view(both_death)
```
Merge all three reports via rbind, call it us_death2
```{r}
hom_suic <- rbind(hom, suic)
#view(hom_suic)
```
```{r}
us_death2 <- rbind(hom_suic, both_death)
#view(us_death2)
```
Remove unnecessary columns from us_death2
```{r}
colnames(us_death2)
```
```{r}
us_death2 <- us_death2 %>% 
  select(-"Sex", -"State", -"Ethnicity", -"Age Group", -"First Year", -"Last Year", -"Population", -"Age-Adjusted Rate")
```
```{r}
us_death2 <- us_death2 %>% 
  rename(Cause_death = "Cause of Death", Metro_type = "2013 Urbanization (Collapsed) Classification", Crude_rate = "Crude Rate", Num_Deaths = "Deaths")
```
```{r}
#view(us_death2)
```
Convert the NAs in Metro_type to All_metros, and the "1" & "2" to Metro & Non Metro
```{r}
us_death2$Metro_type[is.na(us_death2$Metro_type)] <- "All Metros"

```
```{r}
us_death2$Metro_type[us_death2$Metro_type == 1] <- "Metro"
```
```{r}
us_death2$Metro_type[us_death2$Metro_type == 2] <- "Non Metro"
```
```{r}
us_death2$Cause_death[us_death2$Cause_death == "Homicide/Legal Intervention Injury"] <- "Homicide"
```
```{r}
us_death2$Cause_death[us_death2$Cause_death == "Suicide Injury"] <- "Suicide"
```

```{r}
us_death2$Cause_death[us_death2$Cause_death == "Violence-Related Injury"] <- "All Violent Deaths"
```

```{r}
view(us_death2)
```
Test run cause of death plot for app 
```{r}
write.csv(us_death2,"data/us_death2.csv", row.names = TRUE)
```
```{r}
      us_death2 %>%
        filter(Cause_death == "Suicide") %>%
        #filter(Metro_type == input$metro) %>%
        # filter(hispanic == input$hispanic)
        ggplot(aes(x = Race, y = Crude_rate)) +
        geom_col()
```
Consolidate AIAN Missing/Unidentified and AIAN Homicide for map
```{r}
unid_uncl_AIAN <- read_csv("data/NamUs_Unid_Uncl_AIAN_Female.csv")

```
```{r}
colnames(unid_uncl_AIAN)
```

```{r}
unid_uncl_AIAN <- unid_uncl_AIAN %>% 
  select("Case", "City", "County", "State")
```
```{r}
#view(unid_uncl_AIAN)
```
```{r}
miss_AIAN_2 <- read_csv("data/NamUs_Missing_AIAN_Female.csv") 
  
```
```{r}
#view(miss_AIAN_2)
```

```{r}
miss_AIAN_2 %>% 
  rename(Case = "Case Number")
```


```{r}
miss_AIAN_2 <- miss_AIAN_2 %>% 
select(-"DLC", -"Last Name", -"First Name", -"Missing Age", -"Sex", -"Race / Ethnicity", -"Date Modified" )
```
```{r}
#view(miss_AIAN_2)
```
```{r}
miss_AIAN_2 <- miss_AIAN_2 %>% 
  rename(Case = "Case Number")
```
```{r}
#view(miss_AIAN_2)
```
```{r}
muu <- rbind(miss_AIAN_2, unid_uncl_AIAN)
```
```{r}
view(muu)
```
pull in the WISKARS trend report that maps violent deaths by County
```{r}
#death_map <- read_csv("data/WISQARS_AIAN_viol_suic_nonhisp_allmetros_mapping.csv")
#view(death_map)
```
```{r}
#colnames(death_map)
```

```{r}
# death_map <- death_map %>% 
#   select("State Name", "Deaths")
# ```
# ```{r}
# death_map <- death_map %>% 
#   rename(State = "State Name")
```
Convert states to lower case to comply with chloroplethr data.map requirement
```{r}
#tolower(death_map$State)
```
Remove the 52nd row of the deaths dataset for the map, rename "Deaths" to match layout for muu dataset, and convert the Num_Deaths column to numeric.  This will align with the muu file 
 ```{r}
# death_math <- death_map[-c(52),]
 ```
 ```{r}
# death_map %>% 
#   rename(Num_Deaths = "Deaths")
```
```{r}
# death_map_test <- write.csv(death_map, "data/death_map.csv", row.names = TRUE)

```

```{r}
# death_map <- death_map %>%
#   mutate_at(c(2), as.numeric)
```

```{r}
#death_map$Num_Deaths[is.na(death_map$Num_Deaths)] <- 0 
```


#death_map$Num_Deaths[death_map$Num_Deaths == "--"] <- 0

```{r}
muu %>% count(State) %>% 
  rename(Num_MUU = "n")
 
```
Rename all state abbreviations to full state name
```{r}
muu$State[muu$State == "AK"] <- "Alaska"
muu$State[muu$State == "AZ"] <- "Arizona"
muu$State[muu$State == "CA"] <- "California"
muu$State[muu$State == "FL"] <- "Florida"
muu$State[muu$State == "HI"] <- "Hawaii"
muu$State[muu$State == "FL"] <- "Florida"
muu$State[muu$State == "IA"] <- "Iowa"
muu$State[muu$State == "ID"] <- "Idaho"
muu$State[muu$State == "ME"] <- "Maine"
muu$State[muu$State == "MI"] <- "Michigan"
muu$State[muu$State == "MN"] <- "Minnesota"
muu$State[muu$State == "MO"] <- "Missouri"
muu$State[muu$State == "MS"] <- "Mississippi"
muu$State[muu$State == "MT"] <- "Montana"
muu$State[muu$State == "NC"] <- "North Carolina"
muu$State[muu$State == "NE"] <- "Nebraska"
muu$State[muu$State == "NM"] <- "New Mexico"
muu$State[muu$State == "NV"] <- "Nevada"
muu$State[muu$State == "NY"] <- "New York"
muu$State[muu$State == "OK"] <- "Oklahoma"
muu$State[muu$State == "OR"] <- "Oregon"
muu$State[muu$State == "PA"] <- "Pennsylvania"
muu$State[muu$State == "SC"] <- "South Carolina"
muu$State[muu$State == "SD"] <- "South Dakota"
muu$State[muu$State == "TN"] <- "Tennessee"
muu$State[muu$State == "TX"] <- "Texas"
muu$State[muu$State == "UT"] <- "Utah"
muu$State[muu$State == "WA"] <- "Washington"
muu$State[muu$State == "WI"] <- "Wisconsin"

```
```{r}
view(muu)
```
```{r}
muu_state <- muu %>% count(State)
```
```{r}
write.csv(muu_state, "data/muu_state.csv", row.names = TRUE)
```
```{r}

```

```{r}
tolower(muu$State)
```
```{r}
view(muu)
```


read in death_map2 and start over 
```{r}
death_map2 <- read_csv("data/death_map2.csv")
view(death_map2)
```
```{r}
  tolower(death_map2$State)
```
```{r}
death_map2 <- death_map2 %>% 
  select(-"X1")
```

```{r}
death_map2$Deaths[is.na(death_map2$Deaths)] <- 0
```
(above) Remove NAs from death_map2 Deaths column & convert to integer. 
```{r}
death_map2 <- death_map2 %>%
   mutate_at(c(2), as.numeric)
```
```{r}
death_map2 <- death_map2 %>% 
  rename(Num_Deaths = "Deaths")
```
```{r}
#view(death_map2)
```
```{r}

```
```{r}
view(death_map2)
```
```{r}
death_map2 <- death_map2 %>% 
  select(-"State")
```


```{r}
#death_muu_map <- inner_join(death_map2, muu, by = "State")
```
Create plot for death trends which allow filter for between death type, metro type, and count vs rate
```{r}
us_death2 %>% 
  filter(Cause_death == "Homicide", Metro_type == "Non Metro") %>%
  ggplot(aes(x = Year, y = Crude_rate, color = Race))+ 
  geom_line()

  
```

